---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cqtc

<!-- badges: start -->
<!-- badges: end -->

This package provides functions to facilitate and streamline concentration-QTc
analysis of clinical ECG data.

## Installation

You can install the development version of cqtc like so:

``` r
devtools::install_github("rstrotmann/cqtc")
```

## Example

The basic building block for further analyses is the cqtc data object. This
package includes sample data for dofetilide and verapamil copied from
*Parkinson, J., Dota, C. & Rekić, D. Practical guide to
concentration-QTc modeling: a hands-on tutorial. J Pharmacokinet Pharmacodyn 52,
43 (2025)* <https://doi.org/10.1007/s10928-025-09981-8>.

The following is a basic example which demonstrates how to create an exploratory
c-QTc plot for the dofetilide data set. For further analysis, including linear
mixed effects modeling, see `vignette("cqtc")`.

```{r example}
library(cqtc)
library(nif)
library(dplyr)

dof <- dofetilide_cqtc %>%
  cqtc_add_baseline("QTCF", baseline_filter = "NTIME == -0.5") %>% 
  add_bl_popmean("BL_QTCF") %>%
  mutate(DPM_BL_QTCF = BL_QTCF - PM_BL_QTCF) %>%
  derive_group_delta("DQTCF") %>% 
  mutate(NTIME = as.factor(NTIME))


dof <- cqtc(
  dofetilide,
  conc_field = "Cplasma",
  baseline_filter = "NTIME == -0.5")
```

## Hear rate correction

```{r}
#| layout-ncol: 2
rr_plot(dof, "QT")
rr_plot(dof, "QTCF")
```

## Assessment of hysteresis

```{r}
cqtc_time_course_plot(dof, "QTCF")
cqtc_hysteresis_plot(dof)
```

## Exploratory c-QTc plot

```{r}
cqtc_ntile_plot(dof, lm = TRUE, loess = TRUE)
```








