---
title: "Concentration-QTc modeling example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{full-modeling-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(cqtc)
```

# INTRODUCTION

This is a basic tutorial illustrating the use of the cqtc package to conduct
exposure-response analyses for the QT interval of the electrocardiogram.

## Background

Investigation of the pro-arrhythmic potential of new drugs is an explicit
regulatory concern, and the International Council for Harmonisation (ICH) has
issued a dedicated guideline, [ICH
E14](https://www.ema.europa.eu/en/ich-e14-clinical-evaluation-qt-qtc-interval-prolongation-proarrhythmic-potential-non-antiarrhythmic-drugs-scientific-guideline),
that summarizes the expectations to such investigations.

Typically, a dedicated trial to investigate the effect of the drug on the
cardiac repolarization ('thorough QT/QTc study') would satisfy these
expectations. However, alignment between industry partners and regulatory bodies
has been achieved, that a well-designed and conducted QTc assessment based on
concentration-QTc modeling ('c-QTc analysis') can be in many cases sufficient to
exclude clinically relevant QTc effects. This has been summarized in a
scientific white paper by Garnett et al., 2017
(<https://doi.org/10.1007/s10928-017-9558-5>).

The central paradigm of the modeling approach established in this white paper is
to use a pre-specified linear mixed-effects model for the change in the QTc
interval relative to the pre-treatment baseline value ($\Delta QTc$) with fixed
effects of intercept ($\theta _0$), slope ($\theta _2$), influence of baseline
QTc on intercept ($\theta _4$), treatment ($\theta _1$, i.e., active or placebo)
and nominal time from first dose ($\theta _3$). Random effects are included on
the intercept ($\eta _{0,1}$), slope ($\eta _{2,i}$):

$$\Delta QTc_{ijk} = (\theta _0 + \eta _{0,i}) + \theta _1 TRT_j(\theta _2 + \eta _{2,i})C_{ijk} + \theta _3 TIME_j + \theta _4 (QTc_{ij=0} - \overline{QTc_0})$$

Indices refer to subject ($i$), treatment ($j$), and time ($k$).

While this pre-specified model is recommended, alternative approaches can be
used if the data structure or the relationship between drug concentration and
$\Delta QTc$ deviate from the assumptions. For this reason, code for the actual
linear mixed-effects modeling is not part of this package as it may be subject
to specific considerations.

The focus of the cqtc package is to facilitate and standardize the generation of
the c-QTc analysis data set, to support consistency checks and exploratory
analyses, and to check the model-independent assumptions that underlie the
linear mixed-effects modeling approach.

## Outline

This example is based on *Parkinson, J., Dota, C. & Rekić, D. Practical guide to
concentration-QTc modeling: a hands-on tutorial. J Pharmacokinet Pharmacodyn 52,
43 (2025)*. https://doi.org/10.1007/s10928-025-09981-8

The dofetilide and verapamil data sets used in the above publication are
provided as sample data sets as part of this package. The below example makes
use of the dofetilide data.

The first part of this tutorial demonstrates how an analysis-ready c-QTc data set
for dofetilide is constructed from these data. The second part illustrates how
the basic modeling assumptions are checked using exploratory analyses, and the
third part focuses on the actual linear mixed-effects modeling.

The tutorial relies on the following R packages:

```{r setup, warning = F, message=F}
library(dplyr)
library(ggplot2)
library(knitr)
library(lme4)
library(lmerTest)
library(lsmeans)
library(cqtc)
```

# DATA PREPROCESSING

The structure of the dofetilide data set is:

```{r}
head(dofetilide_cqtc, 5)
```

Note that this data set includes data from cross-over treatment with both active
and placebo from the same subjects. This may have consequences on the further
analysis strategy, i.e., it allows modeling of the intra-individual difference
to placebo treatment, $\Delta \Delta QTcF$, rather than $\Delta QTcF$ only.
Real-life data sets may include only data from active treatment, or placebo
data from an independent group of subjects.

For illustrative purposes, and somewhat arbitrarily, we will act in the data
preprocessing part of this tutorial as if the placebo data in the dofetilide
data set were coming from an independent cohort, i.e., not from cross-over
treatment. In the linear mixed-effects modeling part of the tutorial, we will
even discard the data from the placebo treatment entirely, and base the c-QTc
analysis on $\Delta QTc$, i.e., the intra-individual difference to pre-treatment
baseline, because this is a very common analysis type. Refer to the original
publication by Parkinson, et al., 2025 to follow their full (and decidedly more
appropriate) analysis strategy.

To make the dofetilide data accessible for modeling using the aforementioned
pre-specified model, is extended adding the following columns:

* individual baseline QTcF (BL_QTCF):

$$BL\_QTCF = QTcF_{i,0}$$

* the population mean for the baseline QTcF by treatment group
(PM_BL_QTCF):

$$PM\_BL\_QTF = \overline {QTc_0}$$

* the difference between the individual baseline QTcF and the 
respective population mean baseline QTcF:

$$DPM\_BL\_QTCF = QTc_{i,j=0} - \overline {QTc_0}$$

* the difference between DQTCF and the group mean DQTCF of the control
population by time point (DDQTCF):

$$DDQTCF = \Delta \Delta QTcF_{i,t} = \Delta QTcF_{i,t} - \overline{\Delta QTcF_{control,t}}$$

In addition, the numeric nominal time field (NTIME) is changed to a factor
variable:

```{r}
dof <- dofetilide_cqtc %>%
  cqtc_add_baseline("QTCF", baseline_filter = "NTIME == -0.5") |> 
  add_bl_popmean("BL_QTCF") |> 
  mutate(DPM_BL_QTCF = BL_QTCF - PM_BL_QTCF) |> 
  derive_group_delta("DQTCF") |> 
  mutate(NTIME = as.factor(NTIME)) |> 
  cqtc()
```

The last line in the above code pipeline converts the input data to a cqtc
object. This object is essentially only a wrapper around an ordinary R data
frame. However, during the conversion to a cqtc object, some consistency checks
are automatically conducted and missing fields added, where possible. In
addition printing a cqtc object will automatically show only a summary of the
data set:

```{r}
dof
```

The underlying data frame can be printed using:

```{r}
as.data.frame(dof) |> 
  head()
```

# EXPLORATORY DATA ANALYSIS

## Drug effect on heart rate

```{r, fig.width=5, fig.asp=0.7}
rr_plot(dof, "QT", group = "ACTIVE")
rr_plot(dof, "QTCF", group = "ACTIVE")
```

## Assessment of hysteresis

```{r, fig.width=5, fig.asp=0.7}
cqtc_time_course_plot(dof, "QTCF")
```

```{r, fig.width=5, fig.asp=0.7}
cqtc_hysteresis_plot(dof, "DDQTCF")
```

## Linear concentration-QTc relationship

```{r, fig.width=5, fig.asp=0.7}
cqtc_ntile_plot(dof, lm = TRUE, loess = TRUE)
```

# LINEAR MIXED EFFECTS MODELING

## Modeling

```{r, fig.width=5, fig.asp=0.7}
mod <- lmerTest::lmer(
  DQTCF ~ NTIME + ACTIVE + DPM_BL_QTCF + CONC + (CONC||ID),
  data = dof)

# parameter estimates
temp <- as.data.frame(coef(summary(mod, ddf="Kenward-Roger")))
colnames(temp) <- c("estimate", "se", "df", "t", "p")
parameters <- temp %>%
  mutate(
    rse = se/estimate * 100,
      lci = estimate + qt(0.025, df = df) * se,
      uci = estimate + qt(0.975, df = df) * se,
      p = ifelse(p < 0.001, "< 0.001", signif(p, 3))) %>%
  select(estimate, lci, uci, rse, p)

parameters %>% 
  kable(caption = "Model parameters")

grid <- ref.grid(
  mod,
  at = list(
    CONC = seq(0, max(dof$CONC, na.rm = TRUE)),
    ACTIVE = c(FALSE, TRUE),
    DPM_BL_QTCF = 0))

temp1 <- summary(lsmeans::lsmeans(
  grid,
  c("CONC", "ACTIVE"),
  level = 0.9)) %>%
  filter(ACTIVE == TRUE)

dof %>%
  cqtc_ntile_plot(param = "DQTCF", n = 10) +
  geom_line(data = temp1, aes(x = CONC, y = lsmean)) +
  geom_ribbon(
    data = temp1,
    aes(x = CONC, ymin = lower.CL, ymax = upper.CL, y = lsmean),
    alpha = 0.2)
```

## Model diagnostics

```{r, fig.width=5, fig.asp=0.7, results='hide'}
invisible(capture.output(
  cqtc_gof_plot(mod)
))
```





