---
title: "full-modeling-example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{full-modeling-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = F, message=F}
library(tidyverse)
library(knitr)
library(lme4)
library(lmerTest)
library(lsmeans)
library(cqtc)
```

This example is based on Parkinson, J., Dota, C. & Rekić, D. Practical guide to
concentration-QTc modeling: a hands-on tutorial. J Pharmacokinet Pharmacodyn 52,
43 (2025). https://doi.org/10.1007/s10928-025-09981-8

The dofetilide and verapamil data sets used in the publication are provided as
sample data files in this package. This example makes use of the dofetilide data:

```{r}
head(dofetilide_cqtc, 5)
```

# DATA PREPROCESSING

* add individual baseline QTcF (BL_QTCF):

$$DQTCF = \Delta QTcF_{i,t} = QTcF_{i,t} - QTcF_{i,predose}$$

* calculate the population mean for the baseline QTcF by treatment group
(PM_BL_QTCF):

$$PM\_BL\_QTF = \overline {QTc_0}$$

* calculate the difference between the individual baseline QTcF and the 
respective population mean baseline QTcF:

$$DPM\_BL\_QTCF = QTc_{i,j=0} - \overline {QTc_0}$$

* calculate the difference between DQTCF and the group mean DQTCF of the control
population by time point (DDQTCF):

$$DDQTCF = \Delta \Delta QTcF_{i,t} = \Delta QTcF_{i,t} - \overline{\Delta QTcF_{control,t}} $$

* change the numeric nominal time field (NTIME) to factors

```{r}
dof <- dofetilide_cqtc %>%
  mutate(BL_QTCF = QTCF[NTIME == -0.5], .by = c("ID", "ACTIVE")) %>%
  add_bl_popmean("BL_QTCF") %>%
  mutate(DPM_BL_QTCF = BL_QTCF - PM_BL_QTCF) %>%
  derive_group_delta("DQTCF") %>% 
  mutate(NTIME = as.factor(NTIME))
```

# EXPLORATORY DATA ANALYSIS

## Drug effect on heart rate

```{r}
rr_plot(dof, "QT", group = "ACTIVE")
rr_plot(dof, "QTCF", group = "ACTIVE")
```

## Assessment of hysteresis

```{r}
cqtc_time_course_plot(dof, "QTCF")
```

```{r}
cqtc_hysteresis_plot(dof, "DDQTCF")
```

## Linear concentration-QTc relationship

```{r}
cqtc_ntile_plot(dof, lm = TRUE, loess = TRUE)
```

# LINEAR MIXED EFFECTS MODELING

## Modeling

```{r}
mod <- lmerTest::lmer(
  DQTCF ~ NTIME + ACTIVE + DPM_BL_QTCF + CONC + (CONC||ID),
  data = dof)

# parameter estimates
temp <- as.data.frame(coef(summary(mod, ddf="Kenward-Roger")))
colnames(temp) <- c("estimate", "se", "df", "t", "p")
parameters <- temp %>%
  mutate(
    rse = se/estimate * 100,
      lci = estimate + qt(0.025, df = df) * se,
      uci = estimate + qt(0.975, df = df) * se,
      p = ifelse(p < 0.001, "< 0.001", signif(p, 3))) %>%
  select(estimate, lci, uci, rse, p)

parameters %>% 
  kable(caption = "Model parameters")

grid <- ref.grid(
  mod,
  at = list(
    CONC = seq(0, max(dof$CONC, na.rm = TRUE)),
    ACTIVE = c(FALSE, TRUE),
    DPM_BL_QTCF = 0))

temp1 <- summary(lsmeans::lsmeans(
  grid,
  c("CONC", "ACTIVE"),
  level = 0.9)) %>%
  filter(ACTIVE == TRUE)

dof %>%
  cqtc_ntile_plot(param = "DQTCF", n = 10) +
  geom_line(data = temp1, aes(x = CONC, y = lsmean)) +
  geom_ribbon(
    data = temp1,
    aes(x = CONC, ymin = lower.CL, ymax = upper.CL, y = lsmean),
    alpha = 0.2)
```

## Model diagnostics

```{r}
cqtc_gof_plot(mod)
```





